Approximation as Differentiation

### 2025-07-02 main/first-order.tex
Initial pass over AD.
More on lifting, but wondering if the discussion would be better elsewhere..

A programmer, or perhaps a language implementor, might want to embed additional ``approximation points'' into
composite data types, to allow for richer approximation structures. For example, for a list like $1 : 2 : []$
we might want to allow the partial lists $\bot$, $1 : \bot$ and $1 : 2 : \bot$ as approximants, in addition to
the approximants obtained by replacing elements with $\bot$. This can be accomplished by extending the syntax
with a \emph{lifting} type constructor $\tyLift$. Then, as an alternative to the usual definition of lists
of natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could (within the language) define a
type of ``approximation lists'' $\mu\alpha.\tyLift\,(1 + \tyLift\,\synVar{nat} \times \alpha)$, with
monadic computations over approximated lists computing partial lists like $1 : \bot$. Similarly, one could
define a custom approximating version of multiplication $*_{\text{appr}}: \tyLift\,\synVar{nat} \times
\tyLift\,\synVar{nat} \to \tyLift\,\synVar{nat}$, as:
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' \equiv
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\]

\noindent This would allow the programmer to emulate the short-circuiting behaviour of $\mathrm{strictOr}$
from \exref{strict-short-circuit}, or the short-circuiting multiplication used in \citet{perera22} for slicing
matrix convolutions, without baking specific choices of approximation into the interpretation of individual
operators.

With this lifting operator in the core language, one could then choose to fix certain choices of
approximation, by providing a surface language without the lifting monad and interpreting it into the core
language with lifting. The Galois slicing implementations discussed in \citet{perera12a} and
\citet{ricciotti17}, for example, provide an approximation point at every composite type constructor; we can
emulate this in our approach by defining a surface language where the data type of lists $\mu\alpha.1 +
\synVar{nat} \times \alpha$ desugars automatically into the approximating version $\mu\alpha.\tyLift\,(1 +
\tyLift\,\synVar{nat} \times \alpha)$.

@@

\subsection{First-Order Automatic Differentiation via CHAD}

>>>>

A programmer, or perhaps a language implementor, might want to embed additional ``approximation points'' into
composite data types, to allow for richer approximation structures. For example, for a list like $1 : 2 : []$
we might want to allow the partial lists $\bot$, $1 : \bot$ and $1 : 2 : \bot$ as approximants, in addition to
the approximants obtained by replacing elements with $\bot$. This can be accomplished by extending the syntax
with a \emph{lifting} type constructor $\tyLift$. Then, as an alternative to the usual definition of lists of
natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could (within the language) define a type
of ``approximation lists'' $\mu\alpha.\tyLift\,(1 + \tyLift\,\synVar{nat} \times \alpha)$, with monadic
computations over approximated lists computing partial lists like $1 : \bot$. The lattice of approximants of
$1 : 2 : []$ now includes not only the ``shape-preserving'' approximations but also the prefixes obtained by
truncating successive tails with $\bot$:

  \begin{center}
    \begin{tikzpicture}
      \node (top) at (0,0) {$1 : 2 : []$};
      % row 2
      \node [below of=top] (ioi) {$1 : \bot : []$};
      \node [left of=ioi,xshift=-3cm] (oii) {$\bot : 2 : []$};
      \node [right of=ioi,xshift=3cm] (iio) {$1 : 2 : \bot$};
      % row 3
      \node [below of=ioi] (oio) {$\bot : 2 : \bot$};
      \node [left of=oio,xshift=-3cm] (ooi) {$\bot : \bot : []$};
      \node [right of=oio,xshift=3cm] (ioo) {$1 : \bot : \bot$};
      % row 4
      \node [below of=oio] (bot) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};

      % links
      \draw (top) -- (ioi);
      \draw (top) -- (oii);
      \draw (top) -- (iio);
      \draw (ioi) -- (oio);
      \draw (ioi) -- (ooi);
      \draw (ioi) -- (ioo);
    \end{tikzpicture}
  \end{center}

Similarly, one could define a custom approximating version of multiplication $*_{\text{appr}}:
\tyLift\,\synVar{nat} \times \tyLift\,\synVar{nat} \to \tyLift\,\synVar{nat}$, as:
\[x *_{\text{appr}} y = \tmBind{x}{\tmFun{x'}{\syn{if}\;x' \equiv
0\;\syn{then}\;\tmReturn{0}\;\syn{else}\;\tmBind{y}{\tmFun{y'}{\tmReturn{x' * y'}}}}}\]

\noindent This would allow the programmer to emulate the short-circuiting behaviour of $\mathrm{strictOr}$
from \exref{strict-short-circuit}, or the short-circuiting multiplication used in \citet{perera22} for slicing
matrix convolutions, without having to bake specific choices of approximation into the interpretation of
individual operators.

With this lifting operator in the core language, one could then choose to fix certain choices of
approximation, providing a surface language without the lifting monad and interpreting it into the core
language with lifting. The Galois slicing implementations discussed in \citet{perera12a} and
\citet{ricciotti17}, for example, provide an approximation point at every composite type constructor; we can
emulate this in our approach by defining a surface language where the data type of lists $\mu\alpha.1 +
\synVar{nat} \times \alpha$ desugars automatically into the approximating version $\mu\alpha.\tyLift\,(1 +
\tyLift\,\synVar{nat} \times \alpha)$.

@@

\subsection{First-Order Automatic Differentiation via CHAD}
\label{sec:first-order:autodiff}
